---
title: "Visualising the output of Monte Carlo methods with ggsmc"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualising the output of Monte Carlo methods with ggsmc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

The `ggsmc` package uses `ggplot2` to display the results of importance sampling (IS), sequential Monte Carlo (SMC) or ensemble-based algorithms. Each algorithm outputs a collection of points, usually evolved through a sequence of target distributions, which for IS and SMC are weighted. Some functions may also be used to display output from Markov chain Monte Carlo (MCMC) algorithms.

## Data format

To use this package the algorithm output must be in [tidy format](https://r4ds.had.co.nz/tidy-data.html), where each dimension of each parameter for each particle lies in a distinct row in a data frame. The data `sir_cwna_model` provides an example of valid input to the plotting functions.

```{r sir_data}
library(ggsmc)
data("sir_cwna_model")
head(sir_cwna_model)
```

This data contains the output of a particle filter (PF) applied to a target tracking problem. The state $x=(x_1,x_2)^T$ tracked using the PF is two-dimensional, consisting of the (one-dimensional) position and velocity of the target. The value of the first particle for the first target distribution is given by $(x_1,x_2)^T=(3.0163,4.6682)^T$. In tidy format, this is stored on two rows of the data frame, where the `Dimension` column gives the index of the state `ParameterName`: e.g. in the row where `ParameterName=="x"` and `Dimension==2`, the `Value` column gives the value of $x_2$.

If you are unfamiliar with it, this way of storing data with its many repeated values might seem wasteful. Its strength is that this format can be used consistently across different situations, allowing the use of general purpose packages for processing and, in our case, plotting the data. If your data is in the more standard matrix format for Monte Carlo output (i.e. one parameter vector per row), then you can use the [`pivot_longer` function](https://cran.r-project.org/web/packages/tidyr/vignettes/pivot.html).

The `sir_cwna_model` data contains more columns than are required to use the plotting functions in this package. The columns required by all functions are:

- `Target`, which indexes the target distribution, taking a different integer values for each target.

- `Particle`, which indexes the particles, taking a different integer value for each particle.

- `ParameterName`, which uses a string to name each parameter.

- `Dimension`, which indexes the dimension of the parameter, taking a different integer value for each dimension.

- `Value`, which stores the numerical value for the particle, target, parameter and dimension specified by the other columns.

For the output of an IS or SMC algorithm, we may additionally include a `LogWeight` column to store the log of the (normalised) weight of the particle. If this column is not found in the data, each particle will be assigned an equal weight.

## Histograms and densities

The `plot_histogram` and `plot_density` functions may be used to plot, respectively, a histogram or density of the marginal distribution of one dimension of one parameter. If the `LogWeight` column is present in the data a weighted histogram/density will be used.

For these functions, a `parameter` (string) and `dimension` (integer) argument need to be used. If the `target` variable is set, the function will plot the marginal histogram/density for the specified target, parameter and dimension. If no target is specified, the points for all points for the specified parameter and dimension will be used for the plot. The `plot_histogram` function also takes a `bins` argument, which may be used to specify the number of bins for the histogram (if this argument is not specified, the default value in `ggplot2` is used).

For an example, we look at the 20th target of the `sir_cwna_model` data using both a histogram and a density.

```{r histogram}
plot_histogram(sir_cwna_model, parameter = "x", dimension = 1, target = 20, bins = 20)
```


```{r density}
plot_density(sir_cwna_model, parameter = "x", dimension = 1, target = 20)
```

## Scatter plots

The `plot_scatter` function can we used to creates a scatter plot of the Monte Carlo representation of the joint distribution between two parameters, or two dimensions of the same parameter. We specify the parameter and dimension for the x-axis using the arguments `x_parameter` and `x_dimension` respectively. The `target` variable plays the same role as for `plot_histogram` and `plot_density`. If a `LogWeight` column is present in the data, the size of the points in the scatter plot will be used to represent the particle weights.

We illustrate this plot again on the 20th target of the `sir_cwna_model` data.

```{r scatter}
plot_scatter(sir_cwna_model, x_parameter = "x", x_dimension = 1, y_parameter = "x", y_dimension = 2, target = 20, alpha = 0.5, max_size = 3)
```

Note that in this plot we have used two additional arguments, `alpha` and `max_size`, to adjust the look of the plot. In this example we have very few importnace points, I found the default value of `alpha = 0.1` (the transparency of the points from 0 to 1) to be too low, and the default `max_size = 1` (governing the size of the points) to be too small.

## Genealogies

To more clearly understand the output of an SMC or ensemble-based algorithm it can be useful to visualise the evolution of the particles over time. The function `plot_genealogy` may be used for this purpose, for one dimension of one parameter.

## Line graphs

## Animating plots

## Modifying the graph

Titles, axis limits. use_weights, initial.

Modifying with gg.

## Output from nested Monte Carlo algorithms

## Visualising MCMC output
